---
title: "Test"
author: "Andrea Bello"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Libraries
```{r echo = T, results = 'hide'}
library(tidyverse)
library(ggplot2)
library(ggcorrplot)
library(dplyr)
library(corrplot)
library(gridExtra)
library(grid)
library(GGally)
library(fastDummies)
library(stringr)
library(lmtest) 
library(forecast)
library(DIMORA)
library(fpp2)
# library(kableExtra)
```

# Load the data-set
```{r dataset, results = FALSE, message=FALSE}
db <- read_csv("IT_NORD_60min.csv")
str(db)
```

# Assign the mode to missing values
```{r}
IT_NORD_price_day_ahead_nas <- na.omit(db$IT_NORD_price_day_ahead)
na_vector <- is.na(db$IT_NORD_price_day_ahead)
db$IT_NORD_price_day_ahead[na_vector] <- IT_NORD_price_day_ahead_nas
```

# Create the variable 'nord'
```{r}
nord <- db$IT_NORD_price_day_ahead
str(nord)
```

# Create a variable time 'tt'
```{r}
tt<- 1:NROW(db)
str(tt)
plot(tt, nord, xlab="Time", ylab="Prices")
```
# Compute Auto-Correlation with acf
```{r}
acf(nord)
```

# Fit a linear regression model 
```{r}
fit1 <- lm(nord~ tt)
summary(fit1)
```


# Plot of the model
```{r}
plot(tt, nord, xlab="Time", ylab="Prices")
abline(fit1, col=3)
```

# Check the residuals? are they autocorrelated? Test of DW
```{r}
dwtest(fit1)
```

# Check the residuals
```{r}
resfit1<- residuals(fit1)
plot(resfit1, xlab="Time", ylab="residuals" )
```

# Let us do the same with a linear model for time series, so we transform the data into a 'ts' object
```{r}
nord.ts <- ts(nord, frequency = 4)
ts.plot(nord.ts, type="o")
```

# We fit a linear model with the tslm function
```{r}
fitts<- tslm(nord.ts~trend)
```

# Obviously it gives the same results of the first model
```{r}
summary(fitts)
```

```{r}
dwtest(fitts)
```
