---
title: "Project_non_parametric_model"
author: "Andrea Bello"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Libraries
```{r echo = T, results = 'hide'}
library(tidyverse)
library(ggplot2)
library(ggcorrplot)
library(dplyr)
library(corrplot)
library(gridExtra)
library(grid)
library(GGally)
library(fastDummies)
library(stringr)
library(lmtest) 
library(forecast)
library(DIMORA)
library(fpp2)
# library(kableExtra)
```


# Load and visualize the data-set
```{r data_set}
energy <- read.csv("IT_NORD_60min.csv")
head(energy)
```
# Handle missing values (NAs)
In the dataset some variables have missing values. To address the problem, we should know the amount of NAs and the information of the variables containing them. In some cases, the mode/median/mean imputation can be applied, while in other situations the variable can be removed from the dataset. 
The proportion of missing values for every column is shown below.
```{r NAs}
na_counts = (round(colSums(is.na(energy))/nrow(energy), 4))
na_counts[na_counts>0]
```
```{r NA_0}
# Values of 0.0001 suggests that we can use the median imputation without
# worrying to check for the cor test, while 0.0027 of IT_NORD_price_ahead is
# firstly treated in the same way.
IT_NORD_load_actual_entsoe_transparency_median <- median(IT_NORD_load_actual_entsoe_transparency)
na_vector <- is.na(energy$IT_NORD_load_actual_entsoe_transparency)
energy$IT_NORD_load_actual_entsoe_transparency[na_vector] <- IT_NORD_load_actual_entsoe_transparency_median

IT_NORD_load_forecast_entsoe_transparency_median <- median(IT_NORD_load_forecast_entsoe_transparency)
na_vector <- is.na(energy$IT_NORD_load_forecast_entsoe_transparency)
energy$IT_NORD_load_forecast_entsoe_transparency[na_vector] <- IT_NORD_load_forecast_entsoe_transparency_median

IT_NORD_price_day_ahead_median <- median(IT_NORD_price_day_ahead)
na_vector <- is.na(energy$IT_NORD_price_day_ahead)
energy$IT_NORD_price_day_ahead[na_vector] <- IT_NORD_price_day_ahead_median
```


```{r NA_1}
# Calculate correlation before removing NAs
cor_before <- cor.test(energy$IT_NORD_price_day_ahead, as.integer(energy$IT_NORD_solar_generation_actual), na.omit = TRUE)$estimate
# Display the correlation
print(paste("Correlation before NA removal:", cor_before))

# Assign the median to missing values
IT_NORD_solar_generation_actual_median <- median(IT_NORD_solar_generation_actual)
na_vector <- is.na(energy$IT_NORD_solar_generation_actual)
energy$IT_NORD_solar_generation_actual[na_vector] <- IT_NORD_solar_generation_actual_median

# Calculate correlation after NA removal and replacement
cor_after <- cor.test(energy$IT_NORD_price_day_ahead, as.integer(energy$IT_NORD_solar_generation_actual))$estimate
# Display the correlation after NA removal
print(paste("Correlation after NA removal:", cor_after))


# IT_NORD_wind_onshore_generation_actual
# Calculate correlation before removing NAs
cor_before <- cor.test(energy$IT_NORD_price_day_ahead, as.integer(energy$IT_NORD_wind_onshore_generation_actual), na.omit = TRUE)$estimate
# Display the correlation
print(paste("Correlation before NA removal:", cor_before))

# Assign the median to missing values
IT_NORD_wind_onshore_generation_actual_median <- median(IT_NORD_wind_onshore_generation_actual)
na_vector <- is.na(energy$IT_NORD_wind_onshore_generation_actual)
energy$IT_NORD_wind_onshore_generation_actual[na_vector] <- IT_NORD_wind_onshore_generation_actual_median

# Calculate correlation after NA removal and replacement
cor_after <- cor.test(energy$IT_NORD_price_day_ahead, as.integer(energy$IT_NORD_wind_onshore_generation_actual))$estimate
# Display the correlation after NA removal
print(paste("Correlation after NA removal:", cor_after))
```
The cor test is not actually necessary due to the low percentage of NAs compared
to the amount of fields with values. By using the median, we aim at minimizing the potential influence of outliers on the imputed values.
```{r}
print(energy)
```
# Monthly data
```{r monthly}
# Convert timestamps to datetime
energy$cet_cest_timestamp <- as.POSIXct(energy$cet_cest_timestamp, format="%Y-%m-%dT%H:%M:%S%z")

# Aggregate energy to monthly
monthly_energy <- energy %>%
  mutate(month = floor_date(cet_cest_timestamp, "month")) %>%
  group_by(month) %>%
  summarize(
    total_load = sum(IT_NORD_load_actual_entsoe_transparency, na.rm = TRUE),
    total_load_forecast = sum(IT_NORD_load_forecast_entsoe_transparency, na.rm = TRUE),
    total_solar_generation = sum(IT_NORD_solar_generation_actual, na.rm = TRUE),
    total_wind_generation = sum(IT_NORD_wind_onshore_generation_actual, na.rm = TRUE),
    total_price_day_ahead = sum(IT_NORD_price_day_ahead, na.rm = TRUE),
  )

# View the monthly energy
print(monthly_energy)

```
```{r plot_1}
qqnorm(monthly_energy$total_price_day_ahead, pch = 1, frame = FALSE)
qqline(monthly_energy$total_price_day_ahead, col = "steelblue", lwd = 2)
```
Even if the line is not perfect, it follows a distribution as normal as possible.
```{r}
outliers=boxplot.stats(monthly_energy$total_price_day_ahead)$out
length(outliers)/length(monthly_energy$total_price_day_ahead)
```
In this case the outliers account more than 5%, but because the distribution 
doesn't deviate substantially from the normal distribution, we decided to remove
them.
```{r}
# trimming only values that are more extreme than 2.5 and 97.5 percenteline
lower_bound <- quantile(monthly_energy$total_price_day_ahead, 0.025)
upper_bound <- quantile(monthly_energy$total_price_day_ahead, 0.975)
outlier_ind <- which(monthly_energy$total_price_day_ahead < lower_bound | monthly_energy$total_price_day_ahead > upper_bound)
energy_temp <- monthly_energy[- outlier_ind,]

qqnorm(energy_temp$total_price_day_ahead, pch = 1, frame = FALSE)
qqline(energy_temp$total_price_day_ahead, col = "steelblue", lwd = 2)
```
```{r}
# Backup used for testing
backup = monthly_energy
```

```{r}
#Local regression
library(sm)
x <- 1:70
y <- monthly_energy$total_price_day_ahead
plot(x,y)
sm.regression(x, y, h = 10, add = T)

sm.regression(x, y, h = 2, add = T, ngrid=200, col=2)

```

```{r loess}
#Loess (no library required, default tool of R)
plot(x, y, xlab="month", ylab="price day ahead")
lo1 <- loess.smooth(x,y) 
#default span= 0.75
lines(lo1)
# we try with other smoothing parameters 'span' 
lo2 <- loess.smooth(x, y, span=0.9)
lines(lo2,col=2)
lo3 <- loess.smooth(x, y, span=0.4)
lines(lo3,col=3)
```
```{r splines}
#######Regression splines (cubic splines)
#install.packages("splines")
library(splines)
#
plot(x, y, xlab="month", ylab="price day ahead")
#
#we select and identify the knots 'equispaced'
xi<-seq(min(x), max(x), length=4)

#----- Model (2 internal knots)
m1<-lm(y ~ bs(x, knots=xi[2:(length(xi)-1)], degree=3))

###---- for graphical reasons select 200 points where to evaluate the model
xxx<-seq(min(x),max(x),length=200)

#Make predictions by using the 'xxx' points
fit1<-predict(m1, data.frame(x=xxx))
#########
plot(x,y,xlab="month", ylab="price day ahead")
lines(xxx,fit1,col=2)

######vertical lines to indicate the knots
abline(v=xi[2], lty=3)
abline(v=xi[3], lty=3)
```